---
title: "R4MarineScienceW2"
author: "Kylar Duffy"
format: html
---

# Workshop 2: Advanced data wrangling: Extracting ecological signals from noisy systems

## Exercise 1: Cleaning the messy metadata

```{r}
# Load packages
library(tidyverse)
library(palmerpenguins)

island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
) # generate the metadata

print(island_metadata) # view metadata

clean_metadata <- island_metadata |> # create new object
  mutate(island_name = str_trim(island_name), # remove hidden spaces
         island_name = str_to_title(island_name), #fix capitalisation
         station_install = dmy(station_install)) # convert dates

print(clean_metadata)
```

## Exercise 2: The Relational Join

```{r}
# Combine clean metadata with bioloigcal data
penguins_spatial <- penguins |>
  left_join(clean_metadata, # keep all rows from penguins and find matching data from clean_metadata
    by = join_by(island == island_name) # tells r they have different names
  )

head(penguins_spatial) # view new data table
```

## Exercise 3: The Wide Summary Matrix

```{r}
# Step 1. Remove penguins with missing body mass
penguins_spatial |>
  drop_na(body_mass_g) # keep only rows that have an actual value
# Step 2. Group by species and island
max_mass_table <- penguins_spatial |>
  drop_na(body_mass) |> # keep only rows that have an actual value
  group_by(species, island) |> # consider each species x island combo separately
  summarise(max_body_mass = max(body_mass),
    .groups = "drop") |> 
  pivot_wider(
  names_from = island, # take values from island and turn into column names
  values_from = max_body_mass # put max body mass into columns
) 

max_mass_table # view table
```

# Keystone Exercise: Estuary Fish Survey - Data Rescue Mission

## Phase 1: Ingestion and Decontamination

```{r}
# Load packages
library(tidyverse)
library(readxl)

# Import CSV datasets
metadata <- read_csv(
  here::here("data/workshop2/estuary_metadata.csv"))

sonde_data <- read_csv(
  here::here("data/workshop2/estuary_sonde_data.csv"))

species_dictionary <- read_csv(
  here::here("data/workshop2/species_dictionary.csv"))

catch_sheets <- excel_sheets(
  here::here("data/workshop2/estuary_catch_log.xlsx")) 

catch_sheets # view the sheet names

# Read all catch log tabs and combine them into one data frame
catch_data <- map_dfr( # combine all sheets into one data frame by rows
  catch_sheets,
  ~ read_excel(
    here::here("data/workshop2/estuary_catch_log.xlsx"),
    sheet = .x)) # .x is a placeholder for the sheet name

# String standardisation, site and species name
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(site),
    site = str_replace_all(site, " ", "_"),
    species = str_to_lower(species),
    species = str_replace_all(species, " ", "_"))
unique(catch_data$site) # check that above worked
unique(catch_data$species) # check that above worked

# Clean metadata site names
metadata <- metadata |>
  mutate(
    site_name = str_to_lower(site_name),
    site_name = str_replace_all(site_name, " ", "_"))
unique(metadata$site_name) # check

# Check species dictionary
species_dictionary # names match already so no need to clean

# Temporal parsing
glimpse(sonde_data) # tells us what the data types are for each column, timestamp is character data
head(sonde_data$timestamp) # see format of timestamps

# Parse the timestamp
sonde_data <- sonde_data |>
  mutate(
    timestamp = dmy_hm(timestamp))
glimpse(sonde_data) # check that timestamp is now a date-time object

# Remove hardware errors
sum(sonde_data$turbidity == -999, na.rm = TRUE) # check how many -999 values are in the turbidity column
sonde_data <- sonde_data |>
  mutate(
    turbidity = na_if(turbidity, -999)) # replace -999 with NA
sum(sonde_data$turbidity == -999, na.rm = TRUE) # check if worked
```

## Phase 2: Relational Architecture

```{r}
# Summarise the water sensor data into a daily summary table 
sonde_data <- sonde_data |>
  mutate(
    date = floor_date(timestamp, unit = "day")) # create a day variable

# Create a daily summary table
sonde_daily <- sonde_data |> 
  group_by(site, date) |> # treats each site on each day as a group
  summarise(
    mean_temperature = mean(temperature, na.rm = TRUE), # daily mean temp
    mean_salinity = mean(salinity, na.rm = TRUE), # daily mean salinity
    .groups = "drop") # return finished table without keeping it grouped

# Taxonomic translation
catch_translated <- catch_data |>
  left_join(
    species_dictionary,
    by = c("species" = "common_name")) # add scientific names to the catch data
sum(is.na(catch_translated$scientific_name)) # check
catch_translated <- catch_translated |>
  select(-species) # remove the common name column

# The grand join
master_data <- catch_translated |>
  left_join(
    metadata,
    by = c("site" = "site_name")) # join catch data with spatial metadata
master_data <- master_data |>
  left_join(
    sonde_daily,
    by = c("site", "date")) # join daily water quality
``` 

## Phase 3: Zero-Catch Framework

```{r}
# Check structure of sampling days
master_data |>
  distinct(site, date) |>
  count(site) # tells us how many days of sampling were done at each site 

# Complete all species across every site and sampling day
master_data <- master_data |>
  complete(
    site,
    date,
    scientific_name) # make sure all species appear for every site x day, if a species wasn't recorded complete() will create a missing row
glimpse(master_data) # check new rows were added

master_data |>
  summarise(
    missing_count = sum(is.na(count)),
    missing_zone = sum(is.na(zone)),
    missing_salinity = sum(is.na(mean_salinity))) # confirm NA counts

# Restore the spatial metadata for completed zero-catch rows
master_data <- master_data |>
  select(-lat, -lon, -zone) |>
  left_join(
    metadata,
    by = c("site" = "site_name"))
sum(is.na(master_data$zone)) # check

# Restore water quality data for completed zero-catch rows
master_data <- master_data |>
  select(-mean_temperature, -mean_salinity) |>
  left_join(
    sonde_daily,
    by = c("site", "date"))
sum(is.na(master_data$mean_salinity)) # check

# Replace missing catch values with zero
master_data <- master_data |>
  mutate(
    count = coalesce(count, 0)) # keep the existing count where there is one; if count is NA , use 0 instead
sum(is.na(master_data$count)) # check
```

## Phase 4: Statistical Extraction

```{r}
# Group data by species and estuary zone
summary_data <- master_data |>
  group_by(scientific_name, zone)
summary_data

# Calculate summary statistics for fish catch
summary_data <- master_data |>
  group_by(scientific_name, zone) |>
  summarise(
    mean_count = mean(count, na.rm = TRUE), # average fish catch
    sd_count = sd(count, na.rm = TRUE), # standard deviation
    se_count = sd(count, na.rm = TRUE) / sqrt(n()), # standard error
    .groups = "drop") # removes grouping after the summary is finished
summary_data

# Calculate summary statistics for catch and salinity
summary_data <- master_data |>
  group_by(scientific_name, zone) |>
  summarise(
    mean_count = mean(count, na.rm = TRUE),
    sd_count = sd(count, na.rm = TRUE),
    se_count = sd(count, na.rm = TRUE) / sqrt(n()),
    salinity_mean = mean(mean_salinity, na.rm = TRUE),
    salinity_sd = sd(mean_salinity, na.rm = TRUE),
    salinity_se = sd(mean_salinity, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Polish summary table
summary_table <- summary_data |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)))

# Other checks before visualisation
metadata # check data is spatially ordered from upstream to downstream

master_data <- master_data |>
  mutate(
    zone = factor(
      zone,
      levels = c("Upstream", "Middle", "Downstream", "Marine"))) # set the zone order
levels(master_data$zone) #check
```

## Phase 5: Visual Communication

```{r}
# Make a basic abundance–salinity plot
ggplot(
  master_data,
  aes(x = mean_salinity, y = count)
) +
  geom_point() +
  facet_wrap(~ scientific_name) 

# Add appropriate axis labels and clean theme
ggplot(
  master_data,
  aes(x = mean_salinity, y = count)
) +
  geom_point() +
  facet_wrap(~ scientific_name) +
  labs(
    x = "Mean daily salinity",
    y = "Fish abundance"
  ) +
  theme_minimal()

# Add a trend line
ggplot(
  master_data,
  aes(x = mean_salinity, y = count)
) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, colour = "red") +
  facet_wrap(~ scientific_name) +
  labs(
    x = "Mean daily salinity",
    y = "Fish abundance"
  ) +
  theme_minimal()

# Create italic facet labels
species_labels <- c(
  "Acanthopagrus spp." = "italic(Acanthopagrus~spp.)",
  "Lates calcarifer" = "italic(Lates~calcarifer)",
  "Lutjanus argentimaculatus" = "italic(Lutjanus~argentimaculatus)",
  "Platycephalus spp." = "italic(Platycephalus~spp.)") # create a set of labels for scientific names

# Apply italic labels
fish_salinity_plot <- ggplot(
  master_data,
  aes(x = mean_salinity, y = count)
) +
  geom_point(
    size = 1.5,
    alpha = 0.4
  ) +
  geom_smooth(method = "lm", se = TRUE, colour = "red") +
  facet_wrap(
    ~ scientific_name,
    labeller = as_labeller(species_labels, label_parsed)
  ) +
  labs(
    x = "Mean daily salinity",
    y = "Fish abundance"
  ) +
  theme_minimal()

# Save the clean master dataset
write_csv(
  master_data,
  here::here("outputs/tables", "estuary_master_data.csv"))

# Save the summary table
write_csv(
  summary_table,
  here::here("outputs/tables", "estuary_summary_table.csv"))

# Save the figure
ggsave(
  filename = here::here("outputs/figures", "fish_abundance_salinity.png"),
  plot = fish_salinity_plot,
  width = 10,
  height = 7,
  dpi = 300)
```

